---
title:  "Randomized Controlled Trial (RCT)"
author: "Huihang Liu"
date:   "4/29/2021"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We follow the Rubin causal model here and focus on the estimation of ATE in the following. 

## Average Treatment Effect

The averaged difference between the two potential outcomes over all subjects is of interest by researchers, which is termed as the average treatment effect (ATE),

\tau
= \bar{Y}_{T} - \bar{Y}_{C}
= \frac{1}{n} \sum_{i=1}^n (Y_{T,i} - Y_{C,i}) .

Denote by $\tau_i=Y_{T,i} - Y_{C,i}$ the individual treatment effect (ITE) of the $i$th subject. 



## Estimation

### Method 1: Difference between treated and controlled

The simplest method to estimate ATE is using the difference between the treated group and the controlled group, 

\begin{align}
\widehat{\tau}^{(1)}
= \frac{1}{np} \sum_{i\in\{i \mid T_i = 1\}} y_i - \frac{1}{n(1-p)} \sum_{i\in\{i \mid T_i = 0\}} y_i .
\end{align}


### Method 1’: Simple regression

Regression model: 

\begin{align}
  y_i = \tau (T_i - \rho) + \mu + e_i
  (\#eq:model1)
\end{align}


The simple regression estimator $\widehat{\tau}^{(2)}$ is obtained by running a regression of $Y$ against $T$ in model \@ref(eq:model1), i.e.,  
\begin{align}
  \widehat{\tau}^{(2)} = \frac{\sum\left(T_{i}-\bar{T}\right)\left(y_{i}-\bar{y}\right)}{\sum\left(T_{i}-\bar{T}\right)^{2}} = \frac{\sum\left(T_{i}-\rho\right)y_{i}}{n\rho(1-\rho)}
\end{align}


### Method 3: Covariates-adjustment regression

We use the covariates $X$ to regress $y$ which could help to improve the estimate of ATE. The least square estimator for $\tau$ is as follows:

  \begin{align}
  \widehat{\tau}^{(3)} = \{ \tilde{T}^\prime (\mathbf{I} - \mathbf{P}_X) \tilde{T} \}^{-1} \tilde{T}^\prime (\mathbf{I} - \mathbf{P}_X) y .
  (\#eq:lse)
\end{align}

where $\tilde{T}=T-\rho\in\mathbb{R}^{n\times1}$ and  $\mathbf{P}_X \in \mathbb{R}^{p\times p}$ is the projection matrix of $X$. We can find that $\widehat{\tau}^{(2)}$ is a special case of \@ref(eq:lse) where $X$ only have a column of $1$s


## Examples

### Visualization

Let’s first demonstrate a simple example visually 

```{r,results=TRUE}
# settings
set.seed(1001)
n <- 50; p <- 3; rho=0.4

tau <- 1
beta <- c(1, 1, 0.5)

# data
x <- MASS::mvrnorm(n=n,mu=rep(0,p-1), Sigma=diag(p-1))
x <- cbind(rep(1,n), x)
t <- rep(0,n)
t[sample(1:n, n*rho, replace = FALSE)] <- 1
e <- rnorm(n) * 0

y <- tau*t + x%*%beta + e

# method 1
idx_t <- which(t==1)
idx_c <- which(t==0)
tau_hat_1 <- mean(y[idx_t]) - mean(y[idx_c])

# method 2
x_used <- x[,1]
P_x <- x_used %*% solve(t(x_used)%*%x_used) %*% t(x_used)
t_tilde <- t - rho
tau_hat_2 <- solve(t(t_tilde) %*% (diag(n) - P_x) %*% t_tilde) %*% t(t_tilde) %*% (diag(n) - P_x) %*% y

# method 3
x_used <- x[,1:2]
P_x <- x_used %*% solve(t(x_used)%*%x_used) %*% t(x_used)
t_tilde <- t - rho
tau_hat_3 <- solve(t(t_tilde) %*% (diag(n) - P_x) %*% t_tilde) %*% t(t_tilde) %*% (diag(n) - P_x) %*% y

# plot
plot(t, y, xlim=c(-1,2))
points(0, mean(y[idx_c]), col="red", pch=2, cex=2)
points(1, mean(y[idx_t]), col="red", pch=3, cex=2)

plot(x[idx_t,2], y[idx_t], col="blue", pch=2, xlab="x", ylab="y", xlim=c(min(x[,2]), max(x[,2])), ylim=c(min(y), max(y)))
points(x[idx_c,2], y[idx_c], col=rgb(red=0.5, green=0.5, blue=0.5, alpha = 0.5), pch=1)
x_tmp <- x[idx_t,1:2]
beta_t <- solve(t(x_tmp)%*%x_tmp)%*%t(x_tmp)%*%y[idx_t]
x_min <- min(x[,2])
x_max <- max(x[,2])
lines(c(x_min, x_max), c(beta_t[1] + x_min*beta_t[2], beta_t[1] + x_max*beta_t[2]), col="blue")
x_tmp <- x[idx_c,1:2]
beta_c <- solve(t(x_tmp)%*%x_tmp)%*%t(x_tmp)%*%y[idx_c]
lines(c(x_min, x_max), c(beta_c[1] + x_min*beta_c[2], beta_c[1] + x_max*beta_c[2]), col=rgb(red=0.5, green=0.5, blue=0.5, alpha = 0.5))

points(mean(x[idx_c,2]), mean(y[idx_c]), col="red", pch=2, cex=2)
points(mean(x[idx_t,2]), mean(y[idx_t]), col="red", pch=3, cex=2)

legend("topleft", 
    legend=c("Treated","Control"), 
    col=c("blue",rgb(red=0.5, green=0.5, blue=0.5, alpha = 0.5)),
    lty=c(1,1),
    pch=c(2,1),
    lwd=1)
```